---
output:
  html_document: default
  pdf_document: default
---
Explore and Summarize Loan Data from Prosper by Rajvir Reehal 

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(dplyr)
library(gridExtra)
library(scales)
```

```{r echo=FALSE, Load_the_Data}
### Load the Data

pld <- read.csv("prosperLoanData.csv", sep = ',', stringsAsFactors=FALSE)
```
Dataset Summary
```{r echo=FALSE, Dimestions}
dim(pld)
```
There are 113,937 loans with 81 variables.
```{r echo=FALSE, Range_of_Data}
range(pld$ListingCreationDate)
```

We can see the first loan in the dataset originated on 2015-11-09 and the
latest loan we have in the dataset is from 2014-03-10.

The dataset I will be exploring is loan data from Prosper, a peer-to-peer
lending platform. Borrowers can borrow up to $2,000 to $35,000  from the
marketplace and lenders can base their decision on the borrowers credit
history and loan information. 

I will being my analysis with a short summary of the data and then I will
create univariate plots to understand the structure of individual variables 
I am interested in learning more about.

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Prosper_Score}
ggplot(data = pld, aes(x = ProsperScore))+
  geom_histogram(stat = 'count')+
  ylab("Number of Loans")
```

The Prosper Score risk rankings looks normally distriuted with the mean
close to 6.

I noticed 29,084 rows of data consisted of NA. After looking
further into it, I noticed it these loans originated before before July 2009.
Loans that originated before July 2009 do not have data for ProsperScore,
ProsperRating (alpha), ProsperRating (numeric), EstimatedReturn, 
EstimatedEffectiveYield, and EstimatedLoss.

```{r echo=FALSE, message=FALSE, warning=FALSE, Loan_Origination_Quarter}
pld <- mutate(pld, ordered_quarters = 
                paste(substring(pld$LoanOriginationQuarter, 4, 7), 
                      substring(pld$LoanOriginationQuarter, 1, 2), sep = " "))

ggplot(data = pld, aes(x = ordered_quarters))+
  geom_histogram(stat = 'count')+
  theme(axis.text.x = element_text(angle = 90))+
  xlab("Quarter")+
  ylab("Number of Loans Originated")
```

Since the company was founded in 2005 to Q2 of 2008, the company saw a steady
increase in the number of loans originated. But as most of the developed
world dove deep into the recession, the number of loan originations decreased
significantly untl there were no loans originating in Q1 2009.

```{r echo=FALSE, message=FALSE, warning=FALSE, Loan_Original_Amount}
ggplot(data = pld, aes(x = LoanOriginalAmount))+
  geom_histogram(binwidth = 1000)+
  xlab("Original Loan Amount")+
  ylab("Number of Loans")

summary(pld$LoanOriginalAmount)
```

Most of the loans had a principal amount of less than $10,000 with a mean of
$8,337 and median of $6,500. We can see peaks in the plot at multiples of 
$5,000 up to 30,000.

Let's examine that reasons for taking out these loans.

```{r echo=FALSE, message=FALSE, warning=FALSE, Listing_Categories}
### Listing categorical labels
listcatlabels <- c('Debt Consolidation', 'Home Improvement', 'Business', 'Personal Loan', 
       'Student Use', 'Auto', 'Baby & Adoption', 'Boat', 'Cosmetic Procedure', 
       'Engagement Ring', 'Green Loans', 'Household Expenses', 
       'Large Purchases', 'Medical/Dental', 'Motorcycle', 'RV', 'Taxes',
       'Vacation', 'Wedding Loans', 'Other', 'Not Available')

### Factoring categories
pld$ListingCategory <- factor(pld$ListingCategory..numeric., 
                              levels = c(1:6,8:20,7,0), labels = listcatlabels)

ggplot(data = pld, aes(x = ListingCategory))+
  geom_histogram(stat = 'count')+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Count")+
  xlab("Listing Category")
```

Quite a large portion of loans are taken out for the purpose of consolidating
loans. This could indicate that a large amount of borrowers do not have a
high enough income to pay off their loans quickly, or they aren't responsible
with their finances and take on more debt than they should, or they are at
a point in their life where they are taking on debt (fist car, first home,
paying off student loans) and they're just looking to lower their interest
payments by consolidating. Let's examine this further.

```{r echo=FALSE, message=FALSE, warning=FALSE, Income_Ranges}
### Listing income ranges
incomelevels <- c('$0','$1-24,999', '$25,000-49,999',
                  '$50,000-74,999', '$75,000-99,999', 
                  '$100,000+', 'Not displayed', 'Not employed')

### Factoring income ragnes
pld$IncomeRange <- factor(pld$IncomeRange, levels = incomelevels)

ggplot(data = pld, aes(x = IncomeRange))+
  geom_histogram(stat = 'count')+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Count of Borrowers")+
  xlab("Income Range")
```

Most of the borrowers have incomes inbetween $25,000-$49,999 and 
$50,000-$74,999.

```{r echo=FALSE, message=FALSE, warning=FALSE, Employment_Status}

ggplot(data = pld, aes(x = EmploymentStatus))+
  geom_histogram(stat = 'count')+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Count of Borrowers")+
  xlab("Employment Status")
```

Most of the borrwers have identified themselves as employed, however they
did not povide any more details about the type of employment.

```{r echo=FALSE, message=FALSE, warning=FALSE, Employment_Status_Duraton}

ggplot(data = pld, aes(x = EmploymentStatusDuration))+
  geom_histogram(binwidth = 12)+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Count of Borrowers")+
  xlab("Employment Status Duration")

summary(pld$EmploymentStatusDuration)
```

This graph exhibits a right-skewed disribution as the plot immediately preaks
then tails out as the employment status durration increases.
7625 rows of data were removed when making this plot as that many loans did
not have data for EmploymentStatusDuration.

```{r echo=FALSE, message=FALSE, warning=FALSE, Debt_to_Income}

ggplot(data = pld, aes(x = DebtToIncomeRatio))+
  geom_histogram(binwidth = 0.1)+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Count of Borrowers")+
  xlab("Debt To Income")
```

We have a large outlier which is making it difficult to view the plot.
Let's remove the outlier for a better visualization.

```{r echo=FALSE, message=FALSE, warning=FALSE, Debt_to_Income_Without_Outlier}

ggplot(data = pld, aes(x = DebtToIncomeRatio))+
  geom_histogram(binwidth = 0.05)+
  coord_cartesian(xlim = c(0, 1.5))+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Count of Borrowers")+
  xlab("Debt To Income")

summary(pld$DebtToIncomeRatio)

```

The majority of borrowers have a debt-to-income ratio of less than 0.5.
We can see after zooming in, this plot is slightly right-skewed and there
is a tail on the right. The median ratio is 0.22 and 75% of borrowers
have a ratio of less than 0.32.

```{r echo=FALSE, message=FALSE, warning=FALSE, Estimated_Return}

ggplot(data = pld, aes(x = EstimatedReturn))+
  geom_histogram(binwidth = 0.005)+
  ylab("Count of Loans")
  xlab("Estimated Return")

summary(pld$EstimatedReturn)
```

The plot seems to follow a normal distribution with a mean of 9.6%. Which
is a great return margin. Let's plot how much borrowers are paying for their
loans.

```{r echo=FALSE, message=FALSE, warning=FALSE, Borrower_Rate}

ggplot(data = pld, aes(x = BorrowerRate))+
  geom_histogram(binwidth = 0.01)+
  ylab("Count of Loans")
  xlab("Borrower Rate")

summary(pld$BorrowerRate)
```

The plot seems to follow a normal distributon with a right skew. The mean 
borrowing rate is 19.28%. That's a very high rate to pay every year.

Let's examine how long borrowers are borrowing for. It would be very
interesting to see how much of their income goes towards interest payments.

```{r echo=FALSE, message=FALSE, warning=FALSE, Lender_Yield}

ggplot(data = pld, aes(x = LenderYield))+
  geom_histogram(binwidth = 0.01)+
  ylab("Count of Loans")
  xlab("Lender Yield")

summary(pld$LenderYield)
```

Most of the lender yields are between 5% and 35%. The plot follows a normal
distribution with a right-skew and a peak near 31%.

```{r echo=FALSE, message=FALSE, warning=FALSE, Term}

ggplot(data = pld, aes(x = Term))+
  geom_histogram()+
  ylab("Count of Loans")

summary(pld$Term)
```

All of the loans seems to be offered in a 1-year, 3-year, or 5-year term.
Most of the loans are 3-year loans. This is quite a long time to pay a
mean interest borrowing rate of 19.28%. Over three years, a borrower would
be paying almost 60% of the original loan value in interest payments.

```{r echo=FALSE, message=FALSE, warning=FALSE, Loan_Status}

ggplot(data = pld, aes(x = LoanStatus))+
  geom_histogram(stat = "count")+
  coord_trans(y = 'sqrt')+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Count of Loans")
```

Most of the loans are either completed or are current loans. However, there
is a large amount of charged-off and defaulted loans. Further into the
analysis I want to see if the propsre rating is accurate in determining
if a loan will perform or not.


# Univariate Analysis

### What is the structure of your dataset?

The structure of the dataset consists of 81 variables with 113937
observations. Each observation is a loan and the variables describe the
loans with details of the loan and details of the borrower. The observations
are from 2005 to 2014. The variable types are int, num, and ch.

### What is/are the main feature(s) of interest in your dataset?

The main features of interest in the dataset are the yield and return rates
along with the various amount of loan characteristics. This will allow me to
guess what are the common loan characteristics of loans that become
delinquent and what loans characteristics makes the loan a safe investment.

Another feature which interested me what how I can view through plots the
effects of the recession on the loan market

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The other fetures in the dataset that will help support my investigation
are the current credit line, total inquires, revolving monthly payments,
and bank card utilization variables.

### Did you create any new variables from existing variables in the dataset?

I did not see a need to create new variables at this point of the analysis.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

In the Lender Yield and Borrower Rate plots, there is an unusual spike at 
the 31% and 32% points in the plot, respectively,  which I may try to 
explain as I investigate further.

As for the second question, I factorized categorical variables to order them
in a manner to make the plot readable. 

# Bivariate Plots Section

```{r echo=FALSE, Credit_Utilization_vs_Borrower_Rate}
ggplot(data = pld, aes(y = BorrowerRate, x = BankcardUtilization))+
  geom_point(alpha = 1/25)+
  coord_cartesian(xlim = c(0, 2), ylim = c(0, 0.4))+
  ggtitle("Bank Card Utilization vs Borrower Rate")
```

For this plot, I was expecting the borrower rate to increase as bank card
utilization increased because a high bank card utilization rate may indicate
that the borrower is not able to pay off their credit card frequently and
thus has to pay the high interest rates on credit cards. 

And we can see this trend in the plot after adjusting the alpha. As the 
borrower rate increases, so does the bank card utilization. In the area 
where the borrower rate is high and the bank card utilizaton is low, we can
see there are not many loans. An interesting feature of this plot is that
as bank card utilization increases, the borrower rate becomes more
dispersed. It looks like a skewed-funnel.

Maybe there are relationships between the utilization rate and
other variables that are in play here as the utilization rate increases.

Another interesting feature of this plot is that at the 100% utilization 
rate there seems to be a limit, however there are a small amount of loans
above the 100% utilization rate. This could be because of the new credit
requirements that banks and landers had to follow which the government
introduced after the recession or maybe it could be that once borowers hit
100% of their card utilization, they looked elsewhere for credit.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bank_Utilization_Before_And_After_Recession}

### Create subsets of data to compare pre-recession and post-recession bank card
### utilization.

pldpost20090701 <- subset(pld, pld$ListingCreationDate > "2009-07-01")

pldpre20090701 <- subset(pld, pld$ListingCreationDate < "2009-07-01")

ggplot(data = pldpost20090701, aes(y = BorrowerRate, x = BankcardUtilization))+
  geom_point(alpha = 1/25)+
  coord_cartesian(ylim = c(0, 0.4), xlim = c(0, 2))+
  ggtitle("Bank Card Utilization vs Borrower Rate Post Recession")

ggplot(data = pldpre20090701, aes(y = BorrowerRate, x = BankcardUtilization))+
  geom_point(alpha = 1/25)+
  coord_cartesian(ylim = c(0, 0.4), xlim = c(0, 2))+
  ggtitle("Bank Card Utilization vs Borrower Rate Pre Recession")
```

We can see that the limit at 100% utilization in the post recession plot
a more defined than the pre recession plot. There are many more loans
that have > 100% card utilizaion pre recession than post recession, this 
could be due to more stringent borrowing requirements enacted by Prosper
after the recession.

Let's conduct a correlation test to measure the strength of this
relationship. 

```{r, Corrrelation_test_borrowerrate_vs_bankcardutilization}
cor.test(pld$BankcardUtilization, pld$BorrowerRate, method = "pearson")
```

With a p-value of 2.2e-16, this is much less than the significance level of
alpha = 0.05. We can conclude that bank card utilization and borrower rate
are significantly correlated.

Let's continue investigating what effects the borrower rate.

```{r echo=FALSE, message=FALSE, warning=FALSE, Borrower_Rate_vs_Available_BankCard_Credit}
ggplot(data = pldpre20090701, aes(y = BorrowerRate, x = AvailableBankcardCredit))+
  geom_point(alpha = 1/25)+
  coord_cartesian(ylim = c(0, 0.4), xlim = c(0, 50000))+
  ggtitle("Bank Card Utilization vs Available Bank Card Credit")
```

We can see that most of the loans with borrowers that have a high amount of
available credit have a lower borrower rate as expected.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bank_Card_Utilization_vs_Debt_to_Income}

ggplot(data = pld, aes(y = BankcardUtilization, x = DebtToIncomeRatio ))+
  geom_jitter(alpha = 1/25)+
  geom_smooth()+
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1))+
  xlab("Debt To Income")+
  ylab("Bank Card Utilization")+
  ggtitle("Bank Card Utilization vs Debt to Income")
```

The two variables have a positive relationship at a debt to income of 
approximately 0.4, then seems to fall and rise where there are not a great
number of loans.

```{r echo=FALSE, message=FALSE, warning=FALSE, Borrower_Rate_vs_Credit_Score}

### I created new variable averaging the lower and upper range credit scores.

pld$AverageCreditScore <- 
  (pld$CreditScoreRangeLower+pld$CreditScoreRangeUpper)/2

ggplot(data = pld, aes(y = BorrowerRate, x = AverageCreditScore))+
  geom_jitter(alpha = 1/100)+
  geom_smooth()+
  xlim(450, 900)+
  xlab("Average Credit Score")+
  ylab("Borrower Rate")+
  ggtitle("Borrwer Rate vs Average Credit Score")
```

There seems to be a negative relationship between the two variables.
For the credit scores over 750, the borrower rate is low and the ranges of
borrowing rates is smaller for most of the loans compared to borrowing rates
for borrowers with lower credits scores.

Let's compare the borrower rate with the prosper score to see if there is a 
stronger relationship.

```{r echo=FALSE, message=FALSE, warning=FALSE, Borrower_Rate_vs_Prosper_Score}

ggplot(data = na.omit(pld[c('ProsperScore', 'BorrowerRate')]), 
       aes(y = BorrowerRate, x = as.factor(ProsperScore)))+
  geom_boxplot()+
  xlab("Prosper Score")+
  ylab("Borrower Rate")+
  ggtitle("Borrwer Rate vs Prosper Score")
```

As we can see, the higher the prosper score for a loan, the lower the
borrowing rate is.

But how accurate are prosper scores in determining if a loan will perform or
not?

```{r echo=FALSE, message=FALSE, warning=FALSE, Prosper_Score_Accuracy}

### Categorize loans by performing loans (completed, current, final payment in
### progress), non-performing loans (charged off, defaulted), in arrears and, 
### cancelled.
pld <- mutate(pld, PerformanceStatus = ifelse(LoanStatus %in% 
                                 c('Completed', 'Current', 
                                   'FinalPaymentInProgress'), 0, 
                               ifelse(LoanStatus %in%
                                        c('Chargedoff', 'Defaulted'), 2,
                                      ifelse(LoanStatus %in%
                                               c('Cancelled'), 3, 1))))

### Factor categories
pld$PerformanceStatus <- factor(pld$PerformanceStatus, levels = 0:3,
                        labels = c('Performing', 'In Arrears', 'Defaulted',
                                   'Cancelled'))


ggplot(data = pld, aes(x = ProsperScore, fill = PerformanceStatus))+
  geom_bar(position = 'fill', stat = 'count')+
  coord_cartesian(ylim = c(0, 0.5))+
  ylab("Proportion")
```

Prosper scores 5-11 follow the pattern I was expecting. I am surprised by
the difference in the default rates from prosper score 1 to score 2. And
it's surprising that prosper scores 2-4 have a similar performance of loans
with scores of 6-8.

Another interesting observation is the large amount of defaults for loans
originated before before July 2009.

Let's take a look into the return rates for borrowers for each quarter.
Based on this plot, I expect return rates to be much lower before July 2009
than return rates after July 2009.

```{r echo=FALSE, message=FALSE, warning=FALSE, Lender_Return_By_Quarter}
### Create subset dataframe to summarise and calculate the average lender yield
### by quarter.
lender_yield_by_q <- summarise(group_by(pld, ordered_quarters), 
          lender_yield = mean(LenderYield, na.rm = TRUE))

ggplot(data = lender_yield_by_q, aes(x = ordered_quarters, y = lender_yield))+
  geom_line(group=1)+
  theme(axis.text.x = element_text(angle = 90))+
  scale_y_continuous(labels=percent)+
  xlab("Quarter")
```

We have to keep in mind that the plot graphs the estimated lender yield as
it is calculated at the origination of the loan. Lender yield rose until 
Q4 2010 and started falling.

Next I will calculate the actual return on a loan by subracting losses(if 
any) from the original loan amount and will annulize the return. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Actual_Yield}

### I calculate the net payments for the lender and then used this information
### to calculate the annualized rate of return for completed, current, 
### defaulted, and chargedoff loans. For current loans, I calculated the return
### rate for the past 12 months of loans that have existed and are not in 
### arrears for 12 months or over. This chart plots the average return by loan
### inception date and weighs each loan equally.

### Variable for net payments to lender.
NetPayments <- pld$LP_CustomerPayments+
                 pld$LP_ServiceFees+
                 pld$LP_CollectionFees

### Create new dataframe attribute which provides the yield for each loan.
### Loans in final payment and loans in arrears were not included as they
### represent a small portion of total loans.
pld <- mutate(pld, 
              ActualReturnByLoan = ifelse(LoanStatus == 'Completed',
                                        ((NetPayments/LoanOriginalAmount)^(12/LoanMonthsSinceOrigination))-1,
                                      ifelse(LoanStatus == 'Current',
                                        ((MonthlyLoanPayment*12)/LoanOriginalAmount),
                                      ifelse((LoanStatus == 'Defaulted' | LoanStatus == 'Chargedoff'), ((NetPayments/LoanOriginalAmount)^(12/LoanMonthsSinceOrigination))-1,NA))))

### Create subset dataframe to calculate the average yield by quarter.
ActualReturnByloanDF <- summarise(group_by(pld, ordered_quarters), 
          LenderActualYield = mean(ActualReturnByLoan, na.rm = TRUE))

ggplot(data = ActualReturnByloanDF[ActualReturnByloanDF$ordered_quarters!="2009 Q2",], aes(x = ordered_quarters, y = LenderActualYield))+
  geom_line(group=1)+
  theme(axis.text.x = element_text(angle = 90))+
  xlab("Quarter")
```

By the calculations, Prosper lenders experienced losses from Prosper's
inception and through the recession and provided positive returns after 
prosper was required to register with the SEC. Q2 2009 was quite interesting 
as all loans received $0 in customer payments, therefore it was removed from
the plot to make it more readable. 

For loans that have existed for less than 12 months and are in good standing
their returns were annualized.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The return lenders realized before the recession was very interesting as
their return was lower than after the recession. It was also interesting
to see through the plots how Prosper became more strict in they borrowers
that were approved. Looking at the bank utilization plots we can see there
were less borrowers with less than 100% card utilizaton after the recession.

I was expecting to see stronger relationships in the plots between borrower 
rate and other variables such as debt to income and bank card utilization.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I was really interested at how the actual return lenders realized for loans
that originated before the recession were all losses. It's surprising that
there was still interest in Prosper loans after the overall return on the 
lender funds were at loss.

### What was the strongest relationship you found?
The relstionship between the Prosper score and borrower rate.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
ggplot(data = pld, aes(x = ordered_quarters, fill = PerformanceStatus))+
  geom_bar(stat = 'count')+
  theme(axis.text.x = element_text(angle = 90))+
  xlab("Quarter")
```

There are a greater proportion and number of defaulted loans before the
recession that after the recession. This can be attributed to Prosper's
registration with the SEC.

```{r echo=FALSE, Actual_Yield_vs_Estimated_Yield}

pldprerec <- subset(pld, pld$ListingCreationDate > "2009-07-01")
plot1 <- ggplot(aes(x = LenderYield, y = ActualReturnByLoan), data = pldprerec) +
    geom_point(alpha = 0.5) +
    coord_cartesian(xlim = c(0, 0.35))+
    ggtitle("Actual Yield vs Estimated Yield (Pre-Recession)")

pldpostrec <- subset(pld, pld$ListingCreationDate < "2009-07-01")
plot2 <- ggplot(aes(x = LenderYield, y = ActualReturnByLoan), data = pldpostrec) +
    geom_point(alpha = 0.5) +
    coord_cartesian(xlim = c(0, 0.35))+
    ggtitle("Actual Yield vs Estimated Yield (Post-Recession)")

grid.arrange(plot1, plot2, ncol=1, nrow =2)
```

I created these plots to determine how accurate the lender yield was compared
to the actual return of each loan. I split the graphs with pre-recession
and post-recession loans to see the difference in actual returns after 
Prosper registered with the SEC.

We can see for most loans that the actual return was lower than the predicted
lender yield. We can also see the change in distribution between the two
plots before and after the recession; loans that experienced losses
realized greater losses before the recession than after.

```{r echo=FALSE, AvailableCredit_vs_CreditScore_vs_ProsperScore}
plot3 <- ggplot(data = pldpost20090701, aes(y = AvailableBankcardCredit, x = CreditScoreRangeUpper, color = ProsperScore)) +
    geom_point(alpha = 1, position = "jitter") +
    coord_cartesian(xlim = c(600, 800))+
    ggtitle("Available Card Credit vs Credit Score Upper vs Prosper Score")

plot4 <- ggplot(data = pldpost20090701, aes(y = AvailableBankcardCredit, x = CreditScoreRangeLower, color = ProsperScore)) +
    geom_point(alpha = 1, position = "jitter") +
    coord_cartesian(xlim = c(580, 800))+
    ggtitle("Available Card Credit vs Credit Score Lower vs Prosper Score")

grid.arrange(plot3, plot4, ncol = 1, nrow = 2)
```

This plots the relationship between the available card credit and credit 
score and their effect in calculating ther Prosper Score. As credit score
rises, so does the available bank card credit. And in each column grouping
that was created, the share of higher prospe score loans increases as
the cresit score increases.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

As I saw the realtionship that available bank card credit and credit scores 
had with the borrower rate in the bivariate plots section, it indicated 
these variables were used the the model to create the Prosper Score as this
relationship can be viewed in the plots above.

### Were there any interesting or surprising interactions between features?

The large share of defaulted loans in the quarters before the recession
were very surprising to me as it should've been a red flag to lenders
trying to invest their savings in Prosper loans.

The actual return by loan before and after the recession were interesting 
as well. Prosper tightened their borrower requirements to reduce that number
of defaulted loans and it can be visualized through the first two plots.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
No.
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x= DebtToIncomeRatio, y=LenderYield, color=ProsperScore), 
       data=na.omit(filter(pld, DebtToIncomeRatio < 1))) +
  geom_point(alpha = 1) +
  ggtitle("Lender Yield  vs Debt to Income Ratio vs Prosper Rate") +
  xlab ("Debt To Income Ratio < 1") +
  ylab ("Lender Yield")
```

### Description One
This plots the relationship between the estimated lender yield, debt to 
income and the Prosper score. In the bottom left corner where debt to
income is low, the Prosper score is higher and intuitively, the lender yield
is low. But as the lender yield increases, so does the Prosper score and 
debt to income. However, there are still borrowers with low debt to income
and a high lender, this could be due to other factors that make the borrower
more risky.

### Plot Two
```{r echo=FALSE, Plot_Two}

pld$ActualReturnByLoan.bucket <- with(pld, cut(as.numeric(ActualReturnByLoan), 
                                         c(-1, -0.5, 0, 0.1, 0.2, 0.3, 
                                           0.4, 0.5, 1, 2)))

pld$ActualReturnByLoan.bucket <- factor(pld$ActualReturnByLoan.bucket, 
                                      labels = c("-100% to -50%", "-50% to 0%", "0%-10%",
                                      "10%-20%", "20%-30%", "30%-40%", "40%-50%",
                                      "50%-100%", "100%-200%"))

ggplot(data = pld, aes(x = ProsperScore, fill = ActualReturnByLoan.bucket))+
  geom_bar(position = 'fill', stat = 'count')+
  ylab("Proportion")+
  ggtitle("Actual Return by Prosper Score")
```

### Description Two

This plot shows how the relationship between the Prosper score and actual
yield. At the lower end Prosper scores, these loans are deemed very risky,
you can expect a yield anywhere between a great loss or a great positive 
yield. And this is apparent in the proportion of the return buckets for each
Prosper score. The saying "the greater the risk the greater the return...
or loss" is visible here. At the higher end of the Prosper scores, these 
loans are very safe and are the least likely to default, hence their returns
are low. The patterns of the proportions are just as I would expect the to 
be as the Prosper score changes. As the Prosper score goes higher, the 
proportions of loans with very high loans decreases along wih the proportion
of loans with very low returns. 

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(data = ActualReturnByloanDF[ActualReturnByloanDF$ordered_quarters!="2009 Q2",], aes(x = ordered_quarters, y = LenderActualYield))+
  geom_line(group=1, color = 'coral3')+
  geom_line(data = lender_yield_by_q, aes(y = lender_yield), group = 1, color = 'steelblue')+
  theme(axis.text.x = element_text(angle = 90))+
  xlab("Quarter")+
  ylab("Yield")+
  ggtitle("Actual Yield vs Estimated Yield (by quarter)")
```

### Description Three

For the third graph, I decided to keep it simple and plotted the actual 
yield vs the estimated yield. This is to show the great difference between
them two. I'm sure all of the lenders that experienced losses with Prosper
would be very frsustrated at seeing these results as it's apparent that
Prosper didn't do their due diligence in screening their borrowers until
they were forced to register with the SEC.

------

# Reflection

This dataset was very interesting to analyze. There were some variables that
I expected to have a strong relationship to the estimated yield, but didin't.
This made me realize that there are many other variables to consider when
determining the risk of an individual borrower. However, I did expect to see
a stronger relationship between the credit card utilization and the 
borrower rate as credit card utilization, to me, determines if a borrower
understand how inefficient it is to borrow money from a credit card as they
have very high interest rates. If a borrower has high credit card utilization
then I'd deem them to be a risky borrower as they don't seem to understand
what should be their financial priorities. 

This project made me a lot more comfortable with using R as it forced me to
manipulate the data, create different types of plots, create new variables
to see new insights, and to choose the best way to visualize the data. I
struggles with some of the data manipulation, but searching the web made it 
a lot easier to solve as many other people have documented similar struggles
along with their solutions.

If I had more time with the dataset, I would have tried to create a model
that mimics the way the Prosper score is calculated.